---
title: "Get started"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Each view() writes a self-contained page. On this site the page is copied
# next to the article and shown in an iframe, so the maps below are live:
# pan, zoom, and click a feature for its popup.
show_view <- function(v, height = 480) {
  dir <- file.path(knitr::opts_knit$get("output.dir") %||% ".", "views")
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
  name <- paste0(knitr::opts_current$get("label"), ".html")
  stopifnot(file.copy(v$file, file.path(dir, name), overwrite = TRUE))
  htmltools::tags$iframe(
    src = file.path("views", name), width = "100%", height = height,
    style = "border: 1px solid var(--bs-border-color); border-radius: 4px;",
    loading = "lazy", title = paste("Map:", v$name)
  )
}
`%||%` <- function(x, y) if (is.null(x)) y else x
```

aobview is in early development. This page shows what works today; the
[home page](../index.html) lists what is still being worked out.

## Install

aobview and its core package, aobcore, are installed from GitHub:

```r
# install.packages("remotes")
remotes::install_github("allboa/aobview")
```

Vector data can be any geometry wk can read (sf, wkb, wkt, xy, geos, ...) or a data frame with such a column, an Arrow stream or table (a DuckDB result fetched as Arrow, say) or a gdalraster `GDALVector$fetch()` result; PROJ reprojects it. terra and gdalraster are needed for rasters.

## A first view

`view()` takes vector data (here an sf object) and draws it in a page. aobcore ships a
coastline south of 40S in longitude and latitude; aobview sees that it
lies entirely in the far south and draws it in Antarctic Polar
Stereographic (EPSG:3031), densifying edges first so that parallels curve.

```{r coast}
library(aobview)
coast <- sf::st_read(
  system.file("extdata", "coastline_south_40s.geojson", package = "aobcore"),
  quiet = TRUE
)
v <- view(coast)
```

```{r coast-map, echo = FALSE}
show_view(v)
```

In an interactive session, printing the view opens the page in the IDE's
viewer or the browser. `v$file` is the page and `v$scene` the scene it
draws. The page needs no server, and no network unless it references a
remote COG.

## Colour, legends and popups

`zcol` colours features by an attribute and adds a legend; clicking or
tapping a feature shows its attributes. Here are the CCAMLR statistical
areas (a small, simplified copy shipped with aobview, in EPSG:6932),
coloured by area and drawn in EPSG:3031 with the coastline on top.

```{r ccamlr}
areas <- sf::st_read(
  system.file("extdata", "ccamlr_statistical_areas.geojson", package = "aobview"),
  quiet = TRUE
)
areas$area <- paste0("Area ", substr(areas$GAR_Long_Label, 1, 2))
v <- view(areas, crs = "EPSG:3031", zcol = "area",
          popup = c("GAR_Name", "GAR_Long_Label", "GAR_Start_Date", "GAR_Size")) |>
  view_add(coast, popup = FALSE)
```

```{r ccamlr-map, echo = FALSE}
show_view(v, height = 560)
```

Numeric columns take a continuous palette, or classes with `breaks =`;
`palette` is any `grDevices` palette name.

```{r ccamlr-size}
v <- view(areas, zcol = "GAR_Size", palette = "Viridis")
```

```{r ccamlr-size-map, echo = FALSE}
show_view(v)
```

That view is in the data's own CRS, EPSG:6932, because a projected CRS is
kept as it is.

## Any CRS

`crs =` draws the same data in any projected CRS, given as an authority
code, a PROJ string or WKT. A Lambert azimuthal equal-area view centred
on the pole, with 140E at the top:

```{r laea}
v <- view(coast, crs = "+proj=laea +lat_0=-90 +lon_0=140 +datum=WGS84")
```

```{r laea-map, echo = FALSE}
show_view(v)
```

## Rasters

A terra raster is planned into tiles by aobcore and drawn on meshes in
the view CRS, so it is never resampled in R. A remote Cloud Optimized
GeoTIFF is referenced by URL and the browser fetches its tiles; a local
or in-memory raster is embedded. This one is a small polar field in
EPSG:3031 shipped with aobcore, with the coastline over it: the v1
target of a polar raster with vector overlays, in one call.

```{r raster-ok, include = FALSE}
# Raster views need terra, and gdalraster with a PROJ database that resolves
# EPSG:3031; skip the raster example rather than fail the article.
raster_ok <- requireNamespace("terra", quietly = TRUE) &&
  requireNamespace("gdalraster", quietly = TRUE) &&
  !inherits(try(gdalraster::srs_to_wkt("EPSG:3031"), silent = TRUE), "try-error")
```

```{r raster, eval = raster_ok}
sst <- terra::rast(system.file("extdata", "polar_3031.tif", package = "aobcore"))
v <- view(list(sst = sst, coastline = coast))
```

```{r raster-map, echo = FALSE, eval = raster_ok}
show_view(v)
```

Raster views need gdalraster with a working PROJ database. On macOS,
CRAN's gdalraster binary cannot currently find its `proj.db`; gdalraster
from conda-forge works.

## Where next

- [`view()`](../reference/view.html), [`view_add()`](../reference/view-layers.html)
  and [`view_crs()`](../reference/view_crs.html) in the reference.
- [How it fits together](how-it-fits-together.html): aobcore, the scene
  spec, and where lonboard comes in.
- [Get involved](get-involved.html): what to try, and where to report it.
